---
title: "General Social Survey"
author: "Eileen Dunn"  
date: "Summer 2024"  
output: 
  html_document:
    number_sections: true
    toc: true
---


```{r include=FALSE,echo=FALSE}
require(tigerstats)
require(tidyverse)
require(foreign)
source("GrabData.R")
```




#  Question 1

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data in the dataframe GSSdata is from the 2018 General Social Survey. The first blocks of R-code has selected down a subset of the data to just 16 variables. It has further removed unwanted factor levels in much of the data. Examine the code in the GrabData.R file to see what it is doing. Some of the variables are categorical and others are numerical. Be sure to do a variable analysis before tackling each question.  
First question - Is opinion on the death penalty (CAPPUN) independent of gun ownership (OWNGUN)?

$H_0$: Opinion on capital punishment is independent of gun ownership
$H_A$: Opinion on capital punishment is dependent on gun ownership.

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

OWNGUN is categorical, CAPPUN is also categorical 

We want CAPPUN ~ OWNGUN 
The helper file we use is CAT~CAT

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
barchartGC(~OWNGUN + CAPPUN,data=GSSdata)
barchartGC(~OWNGUN + CAPPUN,data=GSSdata, type="percent")

```

This graphs shows that people who own a gun are more likely to favor capital punishment  

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~OWNGUN + CAPPUN,data=GSSdata)
rowPerc(table1)
colPerc(table1)
```

numerical summary shows that 75% of gun owners favor CAPPUN

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

The Chi~squared test p-value is 0.0202, hence we concluded that gun ownership and capital punishment are not independent.   

The fisherexact test has a p-value of .01651 hence we reject the null hypothesis, as with the chi-square test, and furthermore the odds ratio says that gunowners are 2.3 times more likely to favor capital punishment than those who don't own a gun.  

#  Question 2

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Is there a difference in years of education (EDUC) between the those who did or did not vote in 2012 (VOTE12)?

## Introduction
We got this file from GSS.data file.
$H_0$: There is no difference of years EDUC who did VOTE12
$H_A$ There is a difference of years EDUC who did VOTE12

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

EDUC is Num
VOTE12 is a factor
So I used helper file
Num ~ Cat

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
plot(EDUC~VOTE12,data=GSSdata)
```

This boxplot shows that the people who voted in 2012 tend to have a higher education.  

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
favstats(EDUC~VOTE12,data=GSSdata)
```

The Numerical shows that in 2012 those who did vote have almost 16 years in education. Those who did not vote in 2012 have little over 13 years of education.  

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
t.test(EDUC~VOTE12,data=GSSdata)
```

This T tests shows that this means that those who voted in better education.  

#  Question 3

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Determine if a persons performance on the wordsum test (WORDSUM) is independent of their level of schooling (DEGREE).

$H_0$: The score of a persons (WORDSUM) test is independent of their level of schooling (DEGREE)  
$H_A$: The score of a persons (WORDSUM) test is dependent of their level of schooling (DEGREE)  

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

WORDSUM is a numerical variable,
DEGREE is a categorical variable,
so I use Num~Cat data file

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
barchartGC(~WORDSUM + DEGREE,data=GSSdata)
barchartGC(~WORDSUM + DEGREE,data=GSSdata, type="percent")

```

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
table1 <- xtabs(~WORDSUM + DEGREE,data=GSSdata)
rowPerc(table1)
colPerc(table1)
```

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
anova=aov(WORDSUM~DEGREE,data=GSSdata)
summary(anova)
```


This shows that we reject the null because the p-value is 0.0000002 which is smaller than .05

```{r}
plot(TukeyHSD(anova))
```

there is no differnce between highschool and less than highschool. 
there is a difference between Jr college and highschool.
there is a difference with bachelors and highschool
there is a difference between graduate and highschool
No difference between highschool and graduate
No difference between Jr college and graduate
No difference between bachelors and graduate


# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

There is a relationship in all categories.

#### Question 1: is gun ownership independent of opinion on capital punishment. The summary results and the inferential results show that they are not independent.